Formalization of Forcing in Isabelle/ZF \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2020-01-04 IJCAR2020/absoluteness.tex
fuera explicación línea por línea (podría quedar en un apéndice?)

Apart from the axiom schemes, the rest of the $\ZFC$ 
axioms are stated as predicates on classes (that is,
of type \isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); some
of them were already defined by Paulson, and this formulation allows a
better interaction with \isatt{ZF-Constructible}. 
The axioms of Pairing, Union, Foundation,
Extensionality, and Infinity are relativizations of the respective
traditional 
first-order sentences to the class argument. For the Axiom of Choice
we selected a version best suited for the work with transitive
models: the relativization of a sentence stating that for every $x$
there is surjection from an ordinal onto $x$. Finally, Separation and
Replacement were treated separately to effectively obtain first-order
versions. It is to be noted that predicates in Isabelle/ZF are akin to
second order variables and thus do not correspond to first-order
formulas; in particular, there is no induction principle for functions
of type \isatt{i{\isasymRightarrow}o}. For that reason, Separation and
Replacement predicate \emph{on the satisfaction} of a formula $\phi$.
With respect to our previous \cite{2019arXiv190103313G}, we lifted the
arity restriction for the parameter $\phi$ on these schemes,
streamlining various proofs as an immediate benefit and avoiding the
need for tupling.

>>>>

Apart from the axiom schemes, the $\ZFC$ 
axioms are stated as predicates on classes (that is,
of type \isatt{(i{\isasymRightarrow}o){\isasymRightarrow}o}); some
of them were already defined by Paulson, and this formulation allows a
better interaction with \isatt{ZF-Constructible}. 
The axioms of Pairing, Union, Foundation,
Extensionality, and Infinity are relativizations of the respective
traditional 
first-order sentences to the class argument. For the Axiom of Choice
we selected a version best suited for the work with transitive
models: the relativization of a sentence stating that for every $x$
there is surjection from an ordinal onto $x$. Finally, Separation and
Replacement were treated separately to effectively obtain first-order
versions. It is to be noted that predicates in Isabelle/ZF are akin to
second order variables and thus do not correspond to first-order
formulas; in particular, there is no induction principle for functions
of type \isatt{i{\isasymRightarrow}o}. For that reason, Separation and
Replacement predicate \emph{on the satisfaction} of a formula $\phi$.
With respect to our previous \cite{2019arXiv190103313G}, we lifted the
arity restriction for the parameter $\phi$ on these schemes,
streamlining various proofs as an immediate benefit and avoiding the
need for tupling.

### 2020-01-04 IJCAR2020/forcing_theorems.tex
sacando cantidades de líneas
fuera explicación línea por línea (podría quedar en un apéndice?)

As an example of the second kind of induction (on formulas), we choose the
following relatively simple result:

\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ strengthening{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}r{\isasympreceq}p{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env\isanewline 
\ \ \ \ \ {\isasymLongrightarrow}\ r\ {\isasymtturnstile}\ {\isasymphi}\ env{\isachardoublequoteclose}\isanewline
%
%
\isacommand{using}\isamarkupfalse%
\ assms{\isacharparenleft}{\isadigit{2}}{\isacharparenright}
\end{isabelle}
%
The proof is divided in the 4 cases of definition of an element of $\formula$,
%
\begin{isabelle}
\isacommand{proof}\isamarkupfalse%
\ {\isacharparenleft}induct{\isacharparenright}\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Member\ n\ m{\isacharparenright}\isanewline
\ \ \isacommand{then}\isamarkupfalse%
\isanewline
\ \ \dots
\isanewline
\ \ \isacommand{show}\isamarkupfalse%
\ {\isacharquery}case\ \isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ Forces{\isacharunderscore}Member{\isacharbrackleft}of\ {\isacharunderscore}\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ env\ n\ m{\isacharbrackright}\isanewline
\ \ \ \ \ \ strengthening{\isacharunderscore}mem{\isacharbrackleft}of\ p\ r\ {\isachardoublequoteopen}nth{\isacharparenleft}n{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}nth{\isacharparenleft}m{\isacharcomma}env{\isacharparenright}{\isachardoublequoteclose}{\isacharbrackright}\ \isacommand{by}\isamarkupfalse%
\ simp
\end{isabelle}
%
(7 lines omitted) where the final step depends on the previously
proved lemma
\isatt{strengthening{\isacharunderscore}mem} and the characterization of
$\forceisa$ for membership.

The case of equality is entirely analogous, and the \isatt{Nand} and
\isatt{Forall} cases are handled very simply.
%
\begin{isabelle}
\isacommand{next}\isamarkupfalse%
\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Equal\ n\ m{\isacharparenright}\isanewline
\ \ \dots\isanewline
\isacommand{next}\isamarkupfalse%
\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Nand\ {\isasymphi}\ {\isasympsi}{\isacharparenright}\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ assms\isanewline
\ \ \isacommand{show}\isamarkupfalse%
\ {\isacharquery}case\ \isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ Forces{\isacharunderscore}Nand\ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ P{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ pair{\isacharunderscore}in{\isacharunderscore}M{\isacharunderscore}iff\isanewline
\ \ \ \ \ \ Transset{\isacharunderscore}intf{\isacharbrackleft}OF\ trans{\isacharunderscore}M\ {\isacharunderscore}\ leq{\isacharunderscore}in{\isacharunderscore}M{\isacharbrackright}\ leq{\isacharunderscore}transD\ \isacommand{by}\isamarkupfalse%
\ auto\isanewline
\isacommand{next}\isamarkupfalse%
\isanewline
\ \ \isacommand{case}\isamarkupfalse%
\ {\isacharparenleft}Forall\ {\isasymphi}{\isacharparenright}\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ assms\isanewline
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ that\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
\ simp\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ \underline{Forall}\ \isanewline
\ \ \isacommand{have}\isamarkupfalse%
\ {\isachardoublequoteopen}r\ {\isasymtturnstile}\ {\isasymphi}\ {\isacharparenleft}{\isacharbrackleft}x{\isacharbrackright}\ {\isacharat}\ env{\isacharparenright}{\isachardoublequoteclose}\ \isakeyword{if}\ {\isachardoublequoteopen}x{\isasymin}M{\isachardoublequoteclose}\ \isakeyword{for}\ x\isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ that\ pred{\isacharunderscore}le{\isadigit{2}}\ \isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}simp{\isacharparenright}\isanewline
\ \ \isacommand{with}\isamarkupfalse%
\ assms\ \underline{Forall}\isanewline
\ \ \isacommand{show}\isamarkupfalse%
\ {\isacharquery}case\ \isanewline
\ \ \ \ \isacommand{using}\isamarkupfalse%
\ Forces{\isacharunderscore}Forall\ \isacommand{by}\isamarkupfalse%
\ simp\isanewline
\isacommand{qed}\isamarkupfalse
\end{isabelle}
%
It can be noted that the inductive hypothesis
gets used in the last case (underlined here as
\isatt{\underline{Forall}}), but not in the case for \isatt{Nand}.

The main results in this section are the Density and Truth Lemmas,
which we reproduce next.
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ density{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ dense{\isacharunderscore}below{\isacharparenleft}{\isacharbraceleft}q{\isasymin}P{\isachardot}\ {\isacharparenleft}q\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharbraceright}{\isacharcomma}p{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ truth{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ {\isasymlongleftrightarrow}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
%
From these results, the semantical characterization of the forcing
relation (the ``definition of $\forces$''
\cite[IV.2.22]{kunen2011set}) follows easily (49 lines):
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G{\isacharparenright}{\isasymlongrightarrow}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "forcing_in_isabelle_zf"
%%% ispell-local-dictionary: "american"
%%% End:

>>>>

Examples of proofs  using the second kind of induction include
the basic \isatt{strengthening{\isacharunderscore}lemma} and the main
results in this section, the lemmas of  Density  (actually, its nontrivial
direction
\isatt{dense{\isacharunderscore}below{\isacharunderscore}imp{\isacharunderscore}forces})
and Truth, 
which we reproduce next.
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ density{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ dense{\isacharunderscore}below{\isacharparenleft}{\isacharbraceleft}q{\isasymin}P{\isachardot}\ {\isacharparenleft}q\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharbraceright}{\isacharcomma}p{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ truth{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ {\isasymlongleftrightarrow}\ sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
%
From these results, the semantical characterization of the forcing
relation (the ``definition of $\forces$''
\cite[IV.2.22]{kunen2011set}) follows easily:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}{\isacharparenleft}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G{\isacharparenright}{\isasymlongrightarrow}sats{\isacharparenleft}M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}{\isasymphi}{\isacharcomma}map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "forcing_in_isabelle_zf"
%%% ispell-local-dictionary: "american"
%%% End:

### 2020-01-04 IJCAR2020/proper_extension.tex
Choice, y ediciones menores
typos y labels para cada sección
sacando cantidades de líneas
fuera explicación línea por línea (podría quedar en un apéndice?)

\section{Example of proper extension}

@@

Two short theory files contains the relevant
results. \texttt{Ordinals\_In\_MG.thy} shows, using the closure of $M$
under ranks, that $M$ and $M[G]$ share the same ordinals (actually,
ranks of elements of $M[G]$ are bounded by the ranks of their names in
$M$):
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ rank{\isacharunderscore}val{\isacharcolon}\ {\isachardoublequoteopen}rank{\isacharparenleft}val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}{\isacharparenright}\ {\isasymle}\ rank{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\isanewline
\isacommand{lemma}\isamarkupfalse%
\ Ord{\isacharunderscore}MG{\isacharunderscore}iff{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}{\isachardoublequoteclose}\ \isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}

@@

\texttt{Non\_Constructible.thy} contains the first example of a poset
that interprets the locale
\isatt{forcing{\isacharunderscore}notion}. It is the set of all binary
lists with the partial order given by  ``\isatt{xs} is an extension of
\isatt{ys}.'' The sufficient condition for a proper extension is that
the forcing poset is \emph{separative}: every element has two
incompatible extensions.

@@

It is to be noted that both proof-documents comprise mere 99 and 79
(resp.) lines.

>>>>

\section{Example of proper extension}
\label{sec:example-proper-extension}

@@

Two short theory files contain the relevant
results. \texttt{Ordinals\_In\_MG.thy} shows, using the closure of $M$
under ranks, that $M$ and $M[G]$ share the same ordinals (actually,
ranks of elements of $M[G]$ are bounded by the ranks of their names in
$M$):
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ rank{\isacharunderscore}val{\isacharcolon}\ {\isachardoublequoteopen}rank{\isacharparenleft}val{\isacharparenleft}G{\isacharcomma}x{\isacharparenright}{\isacharparenright}\ {\isasymle}\ rank{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\isanewline
\isacommand{lemma}\isamarkupfalse%
\ Ord{\isacharunderscore}MG{\isacharunderscore}iff{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isachardoublequoteopen}Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}{\isachardoublequoteclose}\ \isanewline
\ \ \isakeyword{shows}\ {\isachardoublequoteopen}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}

@@

\texttt{Non\_Constructible.thy} contains our first example of a poset
that interprets the locale
\isatt{forcing{\isacharunderscore}notion}. It is the set of all binary
lists with the partial order given by  ``\isatt{xs} is an extension of
\isatt{ys}.'' The sufficient condition for a proper extension is that
the forcing poset is \emph{separative}: every element has two
incompatible extensions.

@@
